An accelerometer-derived ballistocardiogram method for detecting heartrates in free-ranging marine mammals

### 2021-10-29 analysis/paper/paper.Rmd
Update paper with supplemental references

A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]). Tag configuration and data processing followed the same procedure as the killer whale, with one addition. The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools. Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to manually labeled motionless periods. Motionless periods were identified during or near the bottom phase of dives between fluke strokes. Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].

@@

1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
3. Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
4. Remove noise by applying a triangular moving average smoother.
5. Extract peaks and heuristically remove noisy peaks (see supplemental).

@@

@@

We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. \@ref(fig:bw-profile)).

@@

```{r bw-profile, fig.cap="Blue whale dive profile. Motionless periods indicated by pink boxes."}
ggplot(bw180905_53_10hz, aes(dt, depth)) +
  geom_line(size = 0.5) +
  geom_rect(aes(xmin = start, xmax = stop, ymin = depth_min, ymax = depth_max),
            bw_elg,
            inherit.aes = FALSE,
            fill = "orchid1",
            color = NA,
            alpha = 0.5) +
  scale_x_datetime(date_labels = "%H:%M:%S") +
  scale_y_reverse("Depth (m)") +
  theme_classic() +
  theme(axis.title.x = element_blank())
```

>>>>

A 24.5 m blue whale was tagged with a CATS IMU tag on September 5, 2018 in Monterey Bay, CA under permits MBNMS-MULTI-2017-007, NMFS 21678, and Stanford University IACUC 30123 (previously published in [@goughScalingSwimmingPerformance2019]). Tag configuration and data processing followed the same procedure as the killer whale, with one addition. The 400 Hz acceleration data was used for ballistocardiography (see section **Signal processing**), but we also downsampled the multi-sensor data to 10 Hz for movement analysis using the MATLAB CATS tools.

@@

1. Remove noise and de-trend the acceleration signal with a 5th order Butterworth band-pass filter (killer whale: [1-25Hz], blue whale: [1-10Hz]) [R package `signal` @R-signal].
2. Enhance the IJK complex by differentiating acceleration using a 4th order Savitzky-Golay filter (R package `signal`). Differentiation exaggerates impulses like the J wave.
3. Further enhance the peaks by calculating the Shannon entropy ($H_i=-\sum_{k} |a_{ik}| \times ln(|a_{ik}|)$, where $k$ is the acceleration axis). Additionally, the Shannon entropy is strictly positive, which facilitates peak detection. In the 1d case, $k$ is surge (anterior-posterior acceleration) only.
4. Remove noise by applying a triangular moving average smoother.
5. Extract peaks and heuristically remove noisy peaks (Fig. S1).

@@

Dynamic body movements produce an acceleration signal that masks the ballistocardiogram, so we limited our analyses to motionless periods (Fig. S2). These periods occured during or near the bottom phase of dives between fluke strokes. Strokes were detected from visual examination of the rotational velocity around the lateral axis recorded by gyroscope [_sensu_ @goughScalingSwimmingPerformance2019].

@@

We generated 1d and 3d BCGs for 2 hours of data, including 10 rest dives and `r n_motionless` motionless periods totaling `r format(motionless_minutes, digits = 3)` minutes (Fig. S3).

@@
